Reanalysis Step 3: Propensity Score Analysis with Twang

### 2025-08-17 reanalysis-3-propensity_scores.qmd
twang

---
title: "Reanalysis Step 3: Propensity Score Analysis"
subtitle: "Calculating propensity scores and assessing balance across imputed datasets"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

# Additional packages
library(twang)
library(MatchIt)
library(WeightIt)
library(cobalt)
library(ggplot2)
library(gridExtra)

@@

# Set parameters based on mode
if (analysis_mode == "test") {
  max_imputations_for_diagnostics <- 2  # Show diagnostics for 2 imputations in test
} else {
  max_imputations_for_diagnostics <- 5  # Show diagnostics for up to 5 imputations in final
}

message("Analysis mode: ", analysis_mode)
message("Will show diagnostics for up to ", max_imputations_for_diagnostics, " imputations")
```

@@

## Run Propensity Score Models

We'll use multiple methods for comparison:
1. **Logistic regression** (standard)
2. **GBM via twang** (flexible, handles interactions)
3. **CBPS** (Covariate Balancing Propensity Score)

```{r ps-models}
# Initialize storage for PS results
ps_results <- list()

# Function to calculate PS for one dataset
calculate_ps <- function(data, method = "glm", formula = ps_formula) {

  if (method == "glm") {
    # Standard logistic regression
    ps_model <- glm(formula, data = data, family = binomial())
    ps_scores <- predict(ps_model, type = "response")
    weights <- ifelse(data$ever_lapse_binary == 1, 
                     1, 
                     ps_scores / (1 - ps_scores))

    return(list(
      model = ps_model,
      ps_scores = ps_scores,
      weights = weights,
      method = "glm"
    ))

  } else if (method == "twang") {
    # GBM via twang
    ps_model <- ps(
      formula,
      data = data,
      n.trees = 3000,
      interaction.depth = 3,
      shrinkage = 0.01,
      estimand = "ATT",
      stop.method = c("es.mean", "ks.max"),
      n.minobsinnode = 10,
      verbose = FALSE
    )

    # Get weights for ATT
    weights <- get.weights(ps_model, stop.method = "es.mean")

    return(list(
      model = ps_model,
      ps_scores = ps_model$ps[, "es.mean"],
      weights = weights,
      method = "twang"
    ))

  } else if (method == "cbps") {
    # CBPS via WeightIt
    ps_model <- weightit(
      formula,
      data = data,
      method = "cbps",
      estimand = "ATT"
    )

    return(list(
      model = ps_model,
      ps_scores = ps_model$ps,
      weights = ps_model$weights,
      method = "cbps"
    ))
  }
}

# Calculate PS for each imputed dataset using each method
message("\nCalculating propensity scores...")

for (i in 1:n_imputations) {
  message("Processing imputation ", i, " of ", n_imputations)

  ps_results[[i]] <- list()

  # GLM
  ps_results[[i]][["glm"]] <- calculate_ps(imputed_datasets[[i]], method = "glm")

  # Twang GBM
  ps_results[[i]][["twang"]] <- calculate_ps(imputed_datasets[[i]], method = "twang")

  # CBPS
  ps_results[[i]][["cbps"]] <- calculate_ps(imputed_datasets[[i]], method = "cbps")
}

message("Propensity score calculation complete")
```

@@

```{r balance-assessment}
# Function to assess balance
assess_balance <- function(data, weights, formula = ps_formula) {
  # Create weighted data
  data$weights <- weights

  # Use cobalt for balance assessment
  bal_tab <- bal.tab(
    formula,
    data = data,
    weights = "weights",
    estimand = "ATT",
    stats = c("mean.diffs", "variance.ratios"),
    thresholds = c(m = 0.1, v = 2)
  )

  # Extract key metrics
  balance_summary <- data.frame(
    Mean_SMD = mean(abs(bal_tab$Balance$Diff.Adj), na.rm = TRUE),
    Max_SMD = max(abs(bal_tab$Balance$Diff.Adj), na.rm = TRUE),
    N_Imbalanced = sum(abs(bal_tab$Balance$Diff.Adj) > 0.1, na.rm = TRUE),
    Min_VR = min(bal_tab$Balance$V.Ratio.Adj, na.rm = TRUE),
    Max_VR = max(bal_tab$Balance$V.Ratio.Adj, na.rm = TRUE)
  )

  return(list(
    bal_tab = bal_tab,
    summary = balance_summary
  ))
}

# Assess balance for each method and imputation
balance_results <- list()

for (i in 1:n_imputations) {
  balance_results[[i]] <- list()

  for (method in c("glm", "twang", "cbps")) {
    balance_results[[i]][[method]] <- assess_balance(
      imputed_datasets[[i]],
      ps_results[[i]][[method]]$weights
    )
  }
}

# Summarize balance across imputations
balance_summary_all <- data.frame()

for (method in c("glm", "twang", "cbps")) {
  method_summaries <- lapply(1:n_imputations, function(i) {
    balance_results[[i]][[method]]$summary
  })

  avg_summary <- data.frame(
    Method = method,
    Mean_SMD = mean(sapply(method_summaries, function(x) x$Mean_SMD)),
    Max_SMD = mean(sapply(method_summaries, function(x) x$Max_SMD)),
    N_Imbalanced = mean(sapply(method_summaries, function(x) x$N_Imbalanced)),
    Min_VR = mean(sapply(method_summaries, function(x) x$Min_VR)),
    Max_VR = mean(sapply(method_summaries, function(x) x$Max_VR))
  )

  balance_summary_all <- rbind(balance_summary_all, avg_summary)
}

message("\nBalance Summary Across All Imputations:")
print(balance_summary_all)
kable(balance_summary_all, 
      caption = "Average balance metrics across imputed datasets",
      digits = 3)
```

@@

```{r diagnostics-plots}
#| fig-height: 8
#| fig-width: 12

@@

message("\nGenerating diagnostic plots for ", n_to_show, " imputations...")

# Function to create love plot
create_love_plot <- function(bal_tab_obj, title) {
  love.plot(
    bal_tab_obj,
    stat = "mean.diffs",
    threshold = 0.1,
    title = title,
    colors = c("red", "blue")
  )
}

# Generate plots for each method and selected imputations
for (method in c("glm", "twang", "cbps")) {
  message("\nDiagnostics for ", toupper(method), ":")

  plot_list <- list()

  for (i in imps_to_show) {
    # Love plot
    plot_list[[i]] <- create_love_plot(
      balance_results[[i]][[method]]$bal_tab,
      paste0(toupper(method), " - Imputation ", i)
    )
  }

  # Arrange plots in grid
  if (length(plot_list) > 0) {
    do.call(gridExtra::grid.arrange, c(plot_list, ncol = 2))
  }
}
```

@@

```{r ps-distribution}
#| fig-height: 6
#| fig-width: 10

@@

  # Create data for plotting
  plot_data <- data.frame(
    ps_glm = ps_results[[i]][["glm"]]$ps_scores,
    ps_twang = ps_results[[i]][["twang"]]$ps_scores,
    ps_cbps = ps_results[[i]][["cbps"]]$ps_scores,
    treatment = factor(imputed_datasets[[i]]$ever_lapse_binary,
                      levels = c(0, 1),
                      labels = c("Control", "Treated"))
  )

  # GLM distribution
  p1 <- ggplot(plot_data, aes(x = ps_glm, fill = treatment)) +
    geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
    labs(title = paste("GLM - Imputation", i),
         x = "Propensity Score",
         y = "Count") +
    theme_minimal() +
    scale_fill_manual(values = c("blue", "red"))

  # Twang distribution
  p2 <- ggplot(plot_data, aes(x = ps_twang, fill = treatment)) +
    geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
    labs(title = paste("Twang GBM - Imputation", i),
         x = "Propensity Score",
         y = "Count") +
    theme_minimal() +
    scale_fill_manual(values = c("blue", "red"))

  # CBPS distribution
  p3 <- ggplot(plot_data, aes(x = ps_cbps, fill = treatment)) +
    geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
    labs(title = paste("CBPS - Imputation", i),
         x = "Propensity Score",
         y = "Count") +
    theme_minimal() +
    scale_fill_manual(values = c("blue", "red"))

@@

```{r weight-distribution}
# Summarize weight distributions
weight_summary <- data.frame()

for (method in c("glm", "twang", "cbps")) {
  for (i in 1:n_imputations) {
    weights <- ps_results[[i]][[method]]$weights
    treated <- imputed_datasets[[i]]$ever_lapse_binary == 1

    summary_row <- data.frame(
      Method = method,
      Imputation = i,
      Mean_Weight_Control = mean(weights[!treated]),
      SD_Weight_Control = sd(weights[!treated]),
      Max_Weight_Control = max(weights[!treated]),
      Mean_Weight_Treated = mean(weights[treated]),
      SD_Weight_Treated = sd(weights[treated])
    )

    weight_summary <- rbind(weight_summary, summary_row)
  }
}

# Average across imputations
weight_summary_avg <- weight_summary |>
  group_by(Method) |>
  summarise(
    Mean_Weight_Control = mean(Mean_Weight_Control),
    SD_Weight_Control = mean(SD_Weight_Control),
    Max_Weight_Control = mean(Max_Weight_Control),
    Mean_Weight_Treated = mean(Mean_Weight_Treated),
    SD_Weight_Treated = mean(SD_Weight_Treated)
  )

message("\nWeight Distribution Summary:")
print(weight_summary_avg)
kable(weight_summary_avg,
      caption = "Average weight distributions across imputed datasets",
      digits = 3)
```

@@

```{r save-results}
# Save PS results for use in outcome analysis
ps_output <- list(
  ps_results = ps_results,
  balance_results = balance_results,
  balance_summary = balance_summary_all,
  weight_summary = weight_summary_avg,
  formula_used = ps_formula,
  n_imputations = n_imputations,
  methods_used = c("glm", "twang", "cbps")
)

saveRDS(ps_output, file.path(reanalysis_data_dir, "ps_results.rds"))

message("\nPropensity score results saved to: ", 
        file.path(reanalysis_data_dir, "ps_results.rds"))
```

@@

```{r summary}
# Final summary
message("\n=== PROPENSITY SCORE ANALYSIS SUMMARY ===")
message("Imputations processed: ", n_imputations)
message("Methods used: GLM, Twang GBM, CBPS")
message("Formula included: DR severity (person_dr) and treatment type")

# Best balance
best_method <- balance_summary_all$Method[which.min(balance_summary_all$Mean_SMD)]
message("\nBest balance achieved with: ", best_method)
message("Mean SMD: ", round(balance_summary_all$Mean_SMD[balance_summary_all$Method == best_method], 3))
message("Max SMD: ", round(balance_summary_all$Max_SMD[balance_summary_all$Method == best_method], 3))

# Flag any concerning imbalance
if (any(balance_summary_all$Max_SMD > 0.25)) {
  message("\n⚠ Warning: Some methods have max SMD > 0.25, indicating residual imbalance")
}

if (any(balance_summary_all$Max_VR > 4 | balance_summary_all$Min_VR < 0.25)) {
  message("⚠ Warning: Some methods have extreme variance ratios")
}
```

>>>>

---
title: "Reanalysis Step 3: Propensity Score Analysis with Twang"
subtitle: "Calculating IPTW weights using GBM across imputed datasets"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

# Additional packages
library(twang)
library(survey)
library(cobalt)
library(ggplot2)
library(gridExtra)
library(knitr)

@@

# Set parameters based on mode
if (analysis_mode == "test") {
  max_imputations_for_diagnostics <- 2  # Show diagnostics for 2 imputations in test
  n_trees <- 1000  # Fewer trees for testing
} else {
  max_imputations_for_diagnostics <- 5  # Show diagnostics for up to 5 imputations in final
  n_trees <- 3000  # More trees for final analysis
}

message("Analysis mode: ", analysis_mode)
message("Will show diagnostics for up to ", max_imputations_for_diagnostics, " imputations")
message("Using ", n_trees, " trees for GBM")
```

@@

## Run Twang Propensity Score Models

Following the approach in twang-analysis.Rmd, we'll use GBM via twang package to estimate propensity scores.

```{r ps-models}
# Initialize storage for PS results
twang_results <- list()

# Run twang for each imputed dataset
message("\nCalculating propensity scores using Twang GBM...")
message("Parameters:")
message("  - n.trees: ", n_trees)
message("  - interaction.depth: 3")
message("  - shrinkage: 0.01")
message("  - estimand: ATT")
message("  - stop.method: es.mean, ks.max")

for (i in 1:n_imputations) {
  message("\nProcessing imputation ", i, " of ", n_imputations, "...")

  # Run twang ps model
  twang_model <- ps(
    formula = ps_formula,
    data = imputed_datasets[[i]],
    n.trees = n_trees,
    interaction.depth = 3,
    shrinkage = 0.01,
    estimand = "ATT",
    stop.method = c("es.mean", "ks.max"),
    n.minobsinnode = 10,
    verbose = FALSE
  )

  # Extract weights for ATT (using es.mean as primary stop method)
  weights <- get.weights(twang_model, stop.method = "es.mean")

  # Store results
  twang_results[[i]] <- list(
    model = twang_model,
    weights = weights,
    ps_scores = twang_model$ps[, "es.mean"],
    n_trees_used = twang_model$desc$es.mean$n.trees,
    balance_es_mean = twang_model$balance$es.mean,
    balance_ks_max = twang_model$balance$ks.max
  )

  message("  - Trees used (es.mean): ", twang_model$desc$es.mean$n.trees)
  message("  - Trees used (ks.max): ", twang_model$desc$ks.max$n.trees)
}

message("\nTwang propensity score calculation complete")
```

@@

```{r balance-assessment}
# Extract balance summaries from twang models
balance_summaries <- list()

for (i in 1:n_imputations) {
  # Get balance table from twang
  bal_es <- twang_results[[i]]$balance_es_mean
  bal_ks <- twang_results[[i]]$balance_ks_max

  # Summary statistics for es.mean method
  balance_summaries[[i]] <- data.frame(
    Imputation = i,
    Mean_ES = mean(abs(bal_es$std.eff.sz), na.rm = TRUE),
    Max_ES = max(abs(bal_es$std.eff.sz), na.rm = TRUE),
    N_Imbalanced = sum(abs(bal_es$std.eff.sz) > 0.1, na.rm = TRUE),
    Mean_KS = mean(bal_ks$ks, na.rm = TRUE),
    Max_KS = max(bal_ks$ks, na.rm = TRUE)
  )
}

# Combine all summaries
balance_summary_all <- do.call(rbind, balance_summaries)

# Calculate average across imputations
avg_balance <- data.frame(
  Mean_ES = mean(balance_summary_all$Mean_ES),
  Max_ES = mean(balance_summary_all$Max_ES),
  N_Imbalanced = mean(balance_summary_all$N_Imbalanced),
  Mean_KS = mean(balance_summary_all$Mean_KS),
  Max_KS = mean(balance_summary_all$Max_KS)
)

message("\nBalance Summary Across All Imputations:")
kable(balance_summary_all, 
      caption = "Balance metrics by imputation",
      digits = 3)

message("\nAverage Balance Across Imputations:")
kable(avg_balance,
      caption = "Average balance metrics across all imputations",
      digits = 3)

# Check if balance is acceptable
if (avg_balance$Max_ES > 0.25) {
  message("\n⚠ Warning: Maximum effect size > 0.25, indicating potential residual imbalance")
} else if (avg_balance$Max_ES > 0.1) {
  message("\nNote: Some covariates have effect size > 0.1")
} else {
  message("\n✓ Excellent balance achieved (all effect sizes < 0.1)")
}
```

@@

```{r diagnostics-plots}
#| fig-height: 10
#| fig-width: 12

@@

message("\nGenerating twang diagnostic plots for ", n_to_show, " imputations...")

# For each selected imputation, show twang-specific diagnostics
for (i in imps_to_show) {
  message("\nDiagnostics for Imputation ", i, ":")

  # Get the twang model for this imputation
  twang_model <- twang_results[[i]]$model

  # Plot 1: Balance by iteration (shows convergence)
  plot(twang_model, plots = 1)

  # Plot 2: Standardized effect size by iteration
  plot(twang_model, plots = 2)

  # Plot 3: Balance table plot (if available)
  plot(twang_model, plots = 3)

  # Plot 4: Boxplot of propensity scores by treatment
  plot(twang_model, plots = 4)

  # Additional balance summary plot using twang's summary
  summary(twang_model)
}
```

@@

```{r ps-distribution}
#| fig-height: 8
#| fig-width: 12

@@

  # Create data for plotting
  plot_data <- data.frame(
    ps = twang_results[[i]]$ps_scores,
    treatment = factor(imputed_datasets[[i]]$ever_lapse_binary,
                      levels = c(0, 1),
                      labels = c("Control", "Treated"))
  )

  # Histogram of propensity scores by treatment group
  p1 <- ggplot(plot_data, aes(x = ps, fill = treatment)) +
    geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
    labs(title = paste("Twang GBM Propensity Scores - Imputation", i),
         x = "Propensity Score",
         y = "Count") +
    theme_minimal() +
    scale_fill_manual(values = c("blue", "red")) +
    facet_wrap(~ treatment, ncol = 1, scales = "free_y")

  # Density plot overlay
  p2 <- ggplot(plot_data, aes(x = ps, color = treatment, fill = treatment)) +
    geom_density(alpha = 0.3) +
    labs(title = paste("Density of Propensity Scores - Imputation", i),
         x = "Propensity Score",
         y = "Density") +
    theme_minimal() +
    scale_fill_manual(values = c("blue", "red")) +
    scale_color_manual(values = c("blue", "red"))

  # Boxplot comparison
  p3 <- ggplot(plot_data, aes(x = treatment, y = ps, fill = treatment)) +
    geom_boxplot(alpha = 0.6) +
    geom_jitter(width = 0.1, alpha = 0.2, size = 0.5) +
    labs(title = paste("Distribution Comparison - Imputation", i),
         x = "Treatment Group",
         y = "Propensity Score") +
    theme_minimal() +
    scale_fill_manual(values = c("blue", "red"))

@@

```{r weight-distribution}
# Summarize weight distributions for twang
weight_summary <- data.frame()

for (i in 1:n_imputations) {
  weights <- twang_results[[i]]$weights
  treated <- imputed_datasets[[i]]$ever_lapse_binary == 1

  summary_row <- data.frame(
    Imputation = i,
    Mean_Weight_Control = mean(weights[!treated]),
    SD_Weight_Control = sd(weights[!treated]),
    Min_Weight_Control = min(weights[!treated]),
    Max_Weight_Control = max(weights[!treated]),
    Mean_Weight_Treated = mean(weights[treated]),
    SD_Weight_Treated = sd(weights[treated]),
    Min_Weight_Treated = min(weights[treated]),
    Max_Weight_Treated = max(weights[treated])
  )

  weight_summary <- rbind(weight_summary, summary_row)
}

# Average across imputations
weight_summary_avg <- weight_summary |>
  summarise(
    Mean_Weight_Control = mean(Mean_Weight_Control),
    SD_Weight_Control = mean(SD_Weight_Control),
    Min_Weight_Control = mean(Min_Weight_Control),
    Max_Weight_Control = mean(Max_Weight_Control),
    Mean_Weight_Treated = mean(Mean_Weight_Treated),
    SD_Weight_Treated = mean(SD_Weight_Treated),
    Min_Weight_Treated = mean(Min_Weight_Treated),
    Max_Weight_Treated = mean(Max_Weight_Treated)
  )

message("\nTwang Weight Distribution Summary:")
kable(weight_summary,
      caption = "Weight distributions by imputation",
      digits = 3)

message("\nAverage across all imputations:")
kable(weight_summary_avg,
      caption = "Average weight distributions across imputed datasets",
      digits = 3)

# Check for extreme weights
max_weight <- max(weight_summary$Max_Weight_Control, weight_summary$Max_Weight_Treated)
if (max_weight > 10) {
  message("\n⚠ Warning: Maximum weight exceeds 10, indicating potential instability")
} else if (max_weight > 5) {
  message("\nNote: Some weights exceed 5, monitor for potential instability")
} else {
  message("\n✓ All weights are within reasonable range (<5)")
}
```

@@

```{r save-results}
# Save twang PS results for use in outcome analysis
ps_output <- list(
  twang_results = twang_results,
  balance_summary = balance_summary_all,
  avg_balance = avg_balance,
  weight_summary = weight_summary_avg,
  formula_used = ps_formula,
  n_imputations = n_imputations,
  method = "twang_gbm",
  n_trees = n_trees,
  analysis_mode = analysis_mode
)

saveRDS(ps_output, file.path(reanalysis_data_dir, "ps_results_twang.rds"))

message("\nTwang propensity score results saved to: ", 
        file.path(reanalysis_data_dir, "ps_results_twang.rds"))
```

@@

```{r summary}
# Final summary
message("\n=== TWANG PROPENSITY SCORE ANALYSIS SUMMARY ===")
message("Method: Generalized Boosted Models (GBM) via twang")
message("Imputations processed: ", n_imputations)
message("Analysis mode: ", analysis_mode)
message("Trees used: ", n_trees)
message("Estimand: ATT (Average Treatment Effect on the Treated)")
message("Stop methods: es.mean (effect size), ks.max (Kolmogorov-Smirnov)")

# Enhanced formula summary
message("\nEnhanced formula included:")
message("  - DR severity (person_dr)")
message("  - Treatment type")
message("  - All standard covariates from original analysis")

# Balance summary
message("\nBalance achieved (averaged across imputations):")
message("  - Mean effect size: ", round(avg_balance$Mean_ES, 3))
message("  - Max effect size: ", round(avg_balance$Max_ES, 3))
message("  - Variables with |ES| > 0.1: ", round(avg_balance$N_Imbalanced, 1))
message("  - Mean KS statistic: ", round(avg_balance$Mean_KS, 3))
message("  - Max KS statistic: ", round(avg_balance$Max_KS, 3))

# Balance assessment
if (avg_balance$Max_ES > 0.25) {
  message("\n⚠ Warning: Maximum effect size > 0.25, indicating potential residual imbalance")
  message("  Consider increasing n.trees or adjusting twang parameters")
} else if (avg_balance$Max_ES > 0.1) {
  message("\nNote: Some covariates have effect size > 0.1")
  message("  This is acceptable but monitor in outcome analysis")
} else {
  message("\n✓ Excellent balance achieved (all effect sizes < 0.1)")
}

# Weight stability
message("\nWeight stability:")
message("  - Max weight (control): ", round(weight_summary_avg$Max_Weight_Control, 2))
message("  - Max weight (treated): ", round(weight_summary_avg$Max_Weight_Treated, 2))
```
